---
title: "Milestone 3"
author: "Avinia, Eve, Kelly"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r}
# Import dataset 1, 2 and 3 (Avinia to confirm if dataset 3 OK)

library(tidyverse)
library(dplyr)
library(readr)
library(janitor)

# Read Dataset 1 CSV with the specified arguments
asthma_measures <- readr::read_csv("data/calenviroscreen_measures_2021.csv", 
                            col_names = TRUE, 
                            col_types = NULL, 
                            skip = 0, 
                            skip_empty_rows = TRUE)


# Then clean the names using janitor
asthma_measures <- janitor::clean_names(asthma_measures)

# next work with Dataset 2...
# Read Dataset 2 CSV with the specified arguments
asthma_scores_demog <- readr::read_csv("data/calenviroscreen_scores_demog_2021.csv", 
                            col_names = TRUE, 
                            col_types = NULL, 
                            skip = 0, 
                            skip_empty_rows = TRUE)

# Then clean the names using janitor
asthma_scores_demog <- janitor::clean_names(asthma_scores_demog)

# next work with Dataset 3...
# Read Dataset 3 CSV with the specified arguments
asthma_ER_rate_county <- readr::read_csv("data/asthma-emergency-department-visit-rates-by-county-2015_2022.csv", 
                            col_names = TRUE, 
                            col_types = NULL, 
                            skip = 0, 
                            skip_empty_rows = TRUE)

# Then clean the names using janitor
asthma_ER_rate_county <- janitor::clean_names(asthma_ER_rate_county)

print(asthma_measures);print(asthma_scores_demog);print(asthma_ER_rate_county)

```

## Dataset 1 Code:

# I changed column variables to lower case and underscore and renamed california_county to county for uniformity with other datasets
```{r}

# Change to lower case and underscore, and rename column

asthma_measures_df <- asthma_measures %>% rename_with(~tolower(gsub(" ", "_", .x, fixed = TRUE)))

asthma_measures_df1 <- asthma_measures_df %>% 
  rename(county = california_county)
```

# I selected variables of interest which include asthma ED visits, our primary variable for analysis, four environmental variables and four population characteristics which may be associated with asthma. County and census tract columns were also selected. (11 variables, 8035 obs.)

```{r}

# Select variable of interest

asthma_measures_df2a <- asthma_measures_df1 %>% 
  select(census_tract, county, pm2_5, diesel_pm, 
         traffic, tox_release, asthma, housing_burden, 
         education, unemployment, poverty) 

```

# I removed 11 rows in asthma column with NA values. I did not take out other rows with NAs in order to keep a larger data set to analyze the primary variable asthma. 

```{r}
# Remove rows with NA values in asthma column

asthma_measures_df2a_clean <-
  asthma_measures_df2a[!is.na(asthma_measures_df2a$asthma), ]
                       
                    
```

# I selected median to use as the aggregate measure of CES county level data for asthma ED visits, and its comparison to other population and environmental variables. Median is less influenced by outliers than mean which makes median a valuable method to assess county level distributions. In dataset 1 each county has multiple census level tracts with potential outliers which could distort a more typical county level value. 

```{r}

# Aggregate measure: median asthma rates and environmental variables by county 

asthma_measures_df3a <- asthma_measures_df2a_clean %>% 
  group_by(county) %>% 
  summarize(
    asthma_median_by_county = median(asthma, na.rm = TRUE),
    pm2_5_median_by_county = median(pm2_5, na.rm = TRUE),
    diesel_pm_median_by_county = median(diesel_pm, na.rm = TRUE),
    traffic_median_by_county = median(traffic, na.rm = TRUE),
    tox_release_median_by_county = median(tox_release, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  ungroup()
asthma_measures_df3a <- asthma_measures_df3a %>% 
  arrange(desc(asthma_median_by_county))
```


```{r}

# Aggregate measure: median asthma rates and population characteristics by county

asthma_measures_df3b <- asthma_measures_df2a_clean %>% 
  group_by(county) %>% 
  summarize(
    asthma_median_by_county = median(asthma, na.rm = TRUE),
    housing_burden_median_by_county = median(housing_burden, na.rm = TRUE),
    education_median_by_county = median(education, na.rm = TRUE),
    unemployment_median_by_county = median(unemployment, na.rm = TRUE),
    poverty_median_by_county = median(poverty, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  ungroup()
asthma_measures_df3b <- asthma_measures_df3b %>% 
  arrange(desc(asthma_median_by_county))

```

# Dataset 1 Table Median Asthma Age-Adjusted Rate of ED Visits and Envirnomental Factors by County

```{r}

library(dplyr)
library(kableExtra)

# Median Asthma Rates and Environmental Factors by County

asthma_measures_df3a %>%
  setNames(c("County", "Asthma", "PM 2.5", "Diesel", "Traffic", "Toxic Release")) %>%  # rename columns
  mutate(across(c(Asthma, `PM 2.5`, Diesel, Traffic, `Toxic Release`), ~ round(.x, 3))) %>%  # Round columns to 3 digits
  
  kable(caption = "Median Asthma Age-Adjusted ED Vist Rates and Environmental Factors by County") %>%  kable_styling(
    bootstrap_options = c("striped", "hover"),
    position = "center",
    full_width = FALSE
  ) %>%
  column_spec(1:ncol(asthma_measures_df3a), width = "5em") 
```

# Dataset 1 Table Median Asthma Age-Adjusted ED Visit Rates and Population Characteristics by County

```{r}

asthma_measures_df3b %>%
  setNames(c("County", "Asthma", "Housing Burden", "Education", "Unemployment", "Poverty")) %>%  # rename columns
  mutate(across(c(Asthma, `Housing Burden`, Education, Unemployment, Poverty), ~ round(.x, 3))) %>%  # Round columns to 3 digits
  kable(caption = "Median Asthma Age-Adjusted ED Visit Rates and Population Characteristics by County") %>%  kable_styling(
    bootstrap_options = c("striped", "hover"),
    position = "center",
    full_width = FALSE
  ) %>%
  column_spec(1:ncol(asthma_measures_df3a), width = "5em")  
```

# Scatter plot:  Median Asthma Age-Adjusted ED Visit Rates per County

```{r}
asthma_measures_df13a <- asthma_measures_df3a %>%
mutate(county = factor(county, levels = county[order(asthma_median_by_county)]))

# Plotting

ggplot(asthma_measures_df13a, aes(x = county, y = asthma_median_by_county)) +
  geom_point() +  # Scatter plot points
  labs(
    x = "County",
    y = "Asthma Median",
    title = "Scatter Plot of Median Age-Adjusted ED Visit Rates per County"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)  # Rotate x-axis labels
  )

```
# Dataset 1: Comparative Analysis using Total Counts of Asthma Age-Adjusted ED Vists by County. This shows that county size influences the total ED visit rate which is why median as an aggregating factor may better reflect level of severity of asthma in smaller counties. 

```{r}

asthma_df_counts <- asthma_measures_df2a_clean %>%
  group_by(county) %>%
  summarize(total_count_by_county = sum(asthma))  %>%
  ungroup() %>%
  arrange(desc(total_count_by_county))
```

# Dataset 1 Table: Total Count of Asthma Rate of Age-Adjusted Rate of ED Visis by County

```{r}

asthma_df_counts %>%
  mutate(total_count_by_county = format(total_count_by_county, big.mark = ",", scientific = FALSE)) %>%
  setNames(c("County", "Asthma")) %>%
  kable(caption = "Total Asthma Age-Adjusted Rates of ED Visits by County") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    position = "center",
    full_width = FALSE
  ) %>%
  column_spec(1:2, width = "5em")

```
## Scatter plot:  Total Asthma Age-Adjusted ED Visit Rates per County

```{r}

asthma_df_counts2 <- asthma_df_counts %>%
mutate(county = factor(county, levels = county[order(total_count_by_county)]))

# Plotting

ggplot(asthma_df_counts2, aes(x = county, y = total_count_by_county)) + 
  geom_point() +  # Scatter plot points
  labs(
    x = "County",
    y = "Asthma Total Count",
    title = "Scatter Plot of Total Age-Adjusted ED Visit Rates per County"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)  # Rotate x-axis labels
  )


```



## Dataset 2 Code:

```{r}
# Change to the lower case
asthma_ER_rate_county <- asthma_ER_rate_county %>%
  rename_with(~tolower(gsub(" ", "_", .x, fixed = TRUE)))

# Delete unnecessary column
asthma_ER_rate_county <- asthma_ER_rate_county %>%
  select(-comment)

# Filter the data for the year 2019
asthma_ed_visits_2019 <- asthma_ER_rate_county %>%
  filter(year == 2019)

# Recode Invalid Values
# Replace non-ASCII characters in both age_group and strata_name columns
asthma_ed_visits_2019 <- asthma_ed_visits_2019 %>%
  mutate(
    age_group = str_replace_all(age_group, "[^\x20-\x7E]", "-"),
    strata_name = str_replace_all(strata_name, "[^\x20-\x7E]", "-")
  )

# Check unique values to confirm the cleanup
unique(asthma_ed_visits_2019$age_group)
unique(asthma_ed_visits_2019$strata_name)

# Handle missing data
# Drop rows with NA values
library(tidyverse)

asthma_ed_visits_2019_clean <- asthma_ed_visits_2019 %>%
  drop_na()

str(asthma_ed_visits_2019_clean)
library(tidyverse)
# Descriptive stats per county
descriptive_stats_per_county <- asthma_ed_visits_2019_clean %>%
  group_by(county) %>%
  summarize(
    total_ed_visits = sum(number_of_ed_visits, na.rm = TRUE),
    avg_ed_visits = mean(number_of_ed_visits, na.rm = TRUE),
    median_ed_visits = median(number_of_ed_visits, na.rm = TRUE),
    sd_ed_visits = sd(number_of_ed_visits, na.rm = TRUE),
    min_ed_visits = min(number_of_ed_visits, na.rm = TRUE),
    max_ed_visits = max(number_of_ed_visits, na.rm = TRUE),
    
    avg_age_adjusted_rate = mean(`age_adjusted_ed_visit_rate`, na.rm = TRUE),
    median_age_adjusted_rate = median(`age_adjusted_ed_visit_rate`, na.rm = TRUE),
    sd_age_adjusted_rate = sd(`age_adjusted_ed_visit_rate`, na.rm = TRUE),
    min_age_adjusted_rate = min(`age_adjusted_ed_visit_rate`, na.rm = TRUE),
    max_age_adjusted_rate = max(`age_adjusted_ed_visit_rate`, na.rm = TRUE)
  )
print(descriptive_stats_per_county)


```

Data Dictionary

asthma_median_by_county: median age-adjusted rate of emergency department visits for asthma, by county.

education_median_by_county: median percent of population over 25 with less than a high school education, by county

poverty_median_by_county: median percent of population living below two times the federal poverty level, by county

unemployment_median_by_county: median percent of the population over the age of 16 that is unemployed and eligible for the labor force, by county

housing_burden_median_by_county: median percent housing burdened low income households, by county

pm2_5_median_by_county: median annual mean PM 2.5 concentrations, by county

diesel_pm_median_by_county: median diesel PM emissions from on-road and non-road sources, by county

traffic_median_by_county: median traffic density, in vehicle-kilometers per hour per road length, within 150 meters of the census tract boundary, by county

tox_release_median_by_county: median toxicity-weighted concentrations of modeled chemical releases to air from facility emissions and off-site incineration (from RSEI)






