---
title: "Milestone 4 Joins"
author: "Avinia, Eve, Kelly"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Analysis of California Environmental Factors and Asthma Rates

##The section immediately following is the analysis performed in Milestone 3:
### In this report, we analyze the relationship between **environmental factors**,**asthma rates** and Emergency Department visits across California Counties.
### Each of our 3 datasets has undergone steps to clean and transform the data for analysis: CalEnviro Measures, CalEnviro Scores, and Asthma ER Visits. All datasets are reflecting data from 2019 because CalEnviro Measures and CalEnviro Scores datasets are 2019 year data, so Asthma ER Visits must be filtered to isolate the 2019 yearly data as we prepare for later comparisons across.

##In the later section which is the deliverable for Milestone 4, we perform a final analysiS TO address our 2 research questions:
####1.Compare county asthma ED rates with a county CES measure to assess if there is a correlation. 
####2.Compare asthma ED rates with county-level summaries for specific environmental measures to determine if those specific measures may be worth further investigation.

### Refer to the final section for the Data Dictionary of key variables

```{r}

# Import dataset 1, 2 and 3 (Avinia to confirm if dataset 3 OK)

library(tidyverse)
library(dplyr)
library(readr)
library(janitor)

# Read Dataset 1 CSV with the specified arguments
asthma_measures <- readr::read_csv("data/calenviroscreen_measures_2021.csv", 
                            col_names = TRUE, 
                            col_types = NULL, 
                            skip = 0, 
                            skip_empty_rows = TRUE,
                            na = "NA")

# Then clean the names using janitor
asthma_measures <- janitor::clean_names(asthma_measures)

# next work with Dataset 2...
# Read Dataset 2 CSV with the specified arguments
asthma_scores_demog <- readr::read_csv("data/calenviroscreen_scores_demog_2021.csv", 
                            col_names = TRUE, 
                            col_types = NULL, 
                            skip = 0, 
                            skip_empty_rows = TRUE)

# Then clean the names using janitor
asthma_scores_demog <- janitor::clean_names(asthma_scores_demog)

# next work with Dataset 3...
# Read Dataset 3 CSV with the specified arguments
asthma_ER_rate_county <- readr::read_csv("data/asthma-emergency-department-visit-rates-by-county-2015_2022.csv", 
                            col_names = TRUE, 
                            col_types = NULL, 
                            skip = 0, 
                            skip_empty_rows = TRUE)

# Then clean the names using janitor
asthma_ER_rate_county <- janitor::clean_names(asthma_ER_rate_county)

print(asthma_measures);print(asthma_scores_demog);print(asthma_ER_rate_county)

```


### Dataset 1 CalEnviroScreen Measures: Data Preparation for Analysis

# I changed column variables to lower case and underscore and renamed california_county to county for uniformity with other datasets

```{r}

# Change to lower case and underscore, and rename column

asthma_measures_df <- asthma_measures %>% rename_with(~tolower(gsub(" ", "_", .x, fixed = TRUE)))

asthma_measures_df1 <- asthma_measures_df %>% 
  rename(county = california_county)

```

# I selected variables of interest which include asthma ED visits, our primary variable for analysis, four environmental variables and four population characteristics which may be associated with asthma. County and census tract columns were also selected. (11 variables, 8035 obs.)

```{r}

# selecting variables of interest, (11 variables, 8035 obs.)

asthma_measures_df2a <- asthma_measures_df1 %>% 
  select(census_tract, county, pm2_5, diesel_pm, 
         traffic, tox_release, asthma, housing_burden, 
         education, unemployment, poverty) 

```

# I removed 11 rows in asthma column with NA values. I did not take out other rows with NAs in order to keep a larger data set to analyze the primary variable asthma. 

```{r}

# Remove rows with NA values in asthma column

asthma_measures_df2a_clean <-
  asthma_measures_df2a[!is.na(asthma_measures_df2a$asthma), ]

```

# I selected median to use as the aggregate measure of CES county level data for asthma ED visits, and its comparison to other population and environmental variables. Median is less influenced by outliers than mean which makes median a valuable method to assess county level distributions. In dataset 1 each county has multiple census level tracts with potential outliers which could distort a more typical county level value. 

```{r}
# Aggregate measure: median asthma rates and environmental variables by county 

asthma_measures_df3a <- asthma_measures_df2a_clean %>% 
  group_by(county) %>% 
  summarize(
    asthma_median_by_county = median(asthma, na.rm = TRUE),
    pm2_5_median_by_county = median(pm2_5, na.rm = TRUE),
    diesel_pm_median_by_county = median(diesel_pm, na.rm = TRUE),
    traffic_median_by_county = median(traffic, na.rm = TRUE),
    tox_release_median_by_county = median(tox_release, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  ungroup()
asthma_measures_df3a <- asthma_measures_df3a %>% 
  arrange(desc(asthma_median_by_county))

```

```{r}

# Aggregate measure: median asthma rates and population characteristics by county

asthma_measures_df3b <- asthma_measures_df2a_clean %>% 
  group_by(county) %>% 
  summarize(
    asthma_median_by_county = median(asthma, na.rm = TRUE),
    housing_burden_median_by_county = median(housing_burden, na.rm = TRUE),
    education_median_by_county = median(education, na.rm = TRUE),
    unemployment_median_by_county = median(unemployment, na.rm = TRUE),
    poverty_median_by_county = median(poverty, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  ungroup()
asthma_measures_df3b <- asthma_measures_df3b %>% 
  arrange(desc(asthma_median_by_county))

```

# Dataset 1: Table Median Asthma Age-Adjusted Rate of ED Visits and Envirnomental Factors by County

```{r}

library(dplyr)
library(kableExtra)

# Median Asthma Rates and Environmental Factors by County

asthma_measures_df3a %>%
  setNames(c("County", "Asthma", "PM 2.5", "Diesel", "Traffic", "Toxic Release")) %>%  # rename columns
  mutate(across(c(Asthma, `PM 2.5`, Diesel, Traffic, `Toxic Release`), ~ round(.x, 3))) %>%  # Round columns to 3 digits
  kable(caption = "Median Asthma Age-Adjusted ED Vist Rates and Environmental Factors by County") %>%  kable_styling(
    bootstrap_options = c("striped", "hover"),
    position = "center",
    full_width = TRUE
  ) %>%
  column_spec(1:ncol(asthma_measures_df3a), width = "5em") 
```

# Dataset 1: Table Median Asthma Age-Adjusted ED Visit Rates and Population Characteristics by County

```{r}
# Median Asthma Rates and Population Characteristics by County

asthma_measures_df3b %>%
  setNames(c("County", "Asthma", "Housing Burden", "Education", "Unemployment", "Poverty")) %>%  # rename columns
  mutate(across(c(Asthma, `Housing Burden`, Education, Unemployment, Poverty), ~ round(.x, 3))) %>%  # Round columns to 3 digits
  kable(caption = "Median Asthma Age-Adjusted ED Visit Rates and Population Characteristics by County") %>%  kable_styling(
    bootstrap_options = c("striped", "hover"),
    position = "center",
    full_width = FALSE
  ) %>%
  column_spec(1:ncol(asthma_measures_df3a), width = "5em")  
```


```{r}
# Data for plot Median Asthma Age_adjusted ED Visit Rates per County

asthma_measures_df13a <- asthma_measures_df3a %>%
mutate(county = factor(county, levels = county[order(asthma_median_by_county)]))

# Plotting

ggplot(asthma_measures_df13a, aes(x = county, y = asthma_median_by_county)) +
  geom_point() +  # Scatter plot points
  labs(
    x = "County",
    y = "Asthma Median",
    title = "Scatter Plot of Median Age-Adjusted ED Visit Rates per County"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)  # Rotate x-axis labels
  )

```

# Dataset 1: Comparative Analysis using Total Counts of Asthma Age-Adjusted ED Vists by County. This shows that county size influences the total ED visit rate which is why median as an aggregating factor may better reflect level of severity of asthma in smaller counties. 
```{r}

asthma_df_counts <- asthma_measures_df2a_clean %>%
  group_by(county) %>%
  summarize(total_count_by_county = sum(asthma))  %>%
  ungroup() %>%
  arrange(desc(total_count_by_county))


```

# Dataset 1 Table: Total Count of Asthma Rate of Age-Adjusted Rate of ED Visis by County

```{r}

asthma_df_counts %>%
  mutate(total_count_by_county = format(total_count_by_county, big.mark = ",", scientific = FALSE)) %>%
  setNames(c("County", "Asthma")) %>%
  kable(caption = "Total Asthma Age-Adjusted Rates of ED Visits by County") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    position = "center",
    full_width = TRUE
  ) %>%
  column_spec(1:2, width = "5em")

```

# Scatter plot:  Total Asthma Age-Adjusted ED Visit Rates per County

```{r}
asthma_df_counts2 <- asthma_df_counts %>%
mutate(county = factor(county, levels = county[order(total_count_by_county)]))

# Plotting

ggplot(asthma_df_counts2, aes(x = county, y = total_count_by_county)) + 
  geom_point() +  # Scatter plot points
  labs(
    x = "County",
    y = "Asthma Total Count",
    title = "Scatter Plot of Total Age-Adjusted ED Visit Rates per County"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)  # Rotate x-axis labels
  )

```

### Dataset 2 CalEnviroScreen Scores: Data Preparation for Analysis
#### Selection of Race/Ethnicity as demographic factor of interest for analysis
```{r}
#clean up my race/ethnicity columns
asthma_scores_demog <- asthma_scores_demog %>%
  rename(
    hispanic = hispanic_percent,
    white = white_percent,
    black = african_american_percent,
    native_american = native_american_percent,
    asian = asian_american_percent,
    other = other_multiple_percent
  )

```


```{r}
#check we have values in race/ethnicity and CES score
race_cols <- c("hispanic", "white", "black", "native_american", "asian", "other")

#filter  data based on criteria
asthma_scores_filtered <- asthma_scores_demog %>%
  #keep rows if at least one race column is non-zero/non-NA
  filter(if_any(all_of(race_cols), ~ !is.na(.x) & .x != 0)) %>%
  #keep rows if CES score is non-NA and greater than zero
  filter(!is.na(ces_4_0_score) & ces_4_0_score != 0)
```

#### Subsetting rows and columns for county-level analysis, and creation of new variables: 1) race/ethnicity as weighted avg. of total pop, 2) mean CES
```{r}
#group by county and calculate simple or weighted averages
asthma_scores_county <- asthma_scores_filtered %>%
  group_by(county) %>%
  summarise(
    avg_ces_score = mean(ces_4_0_score),
    
    #population-weighted average for race/ethnicity percentages
    hispanic = sum(hispanic * total_population) / sum(total_population),
    white = sum(white * total_population) / sum(total_population),
    black = sum(black * total_population) / sum(total_population),
    native_american = sum(native_american * total_population) / sum(total_population),
    asian = sum(asian * total_population) / sum(total_population),
    other = sum(other * total_population) / sum(total_population),
    
    #include a total population sum for each county
    total_population = sum(total_population)
  )

#round race/ethnicity percentages to 1 decimal place in preparation for table
asthma_scores_county <- asthma_scores_county %>%
  mutate(
    across(c(hispanic, white, black, native_american, asian, other), ~ round(.x, 1))
  )

```

### Dataset 2 CalEnviroScreen Scores: Table and Plot

```{r}
library(dplyr)
library(knitr)
library(kableExtra)
library(stringr)

#rename CES column header to title case and "CES" in caps
asthma_scores_county <- asthma_scores_county %>%
  rename("mean_CES_score" = "avg_ces_score") %>%  
  rename_with(~ gsub("_", " ", .x)) 

#change the rest of my column headers to Title Case
colnames(asthma_scores_county) <- colnames(asthma_scores_county) %>%
  str_replace_all("_", " ") %>%
  str_to_title() %>%
  str_replace("Ces", "CES")  #manually need to stop reverting lower case
 

#create kable table
asthma_scores_county %>%
  kable(
    caption = "County-Level CalEnviroScore (CES) with Race/Ethnicity",
    format = "html",   
    digits = 2         
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE, 
    position = "center"
  ) %>%
  column_spec(2:8, bold = TRUE)

```

#### Further data preparation for plot of Race/Ethnicity by mean CES score

```{r}

library(dplyr)
library(tidyr)
library(ggplot2)
install.packages("RColorBrewer")
library(RColorBrewer)

#change my data into long format
asthma_scores_plot <- asthma_scores_county %>%
  pivot_longer(
    cols = c("Hispanic", "White", "Black", "Native American", "Asian", "Other"),
    names_to = "Race",
    values_to = "Proportion"
  )

#rank CA counties by mean CES score for plotting
asthma_scores_plot <- asthma_scores_plot %>%
  arrange(desc(`Mean CES Score`)) %>%  
  mutate(County = factor(County, levels = unique(County)))

#remove "County" from the county names in the labels
asthma_scores_plot <- asthma_scores_plot %>%
  mutate(County = str_replace(County, " County", ""))

#make my plot
ggplot(asthma_scores_plot, aes(x = County, y = Proportion, fill = Race)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Proportion of Race/Ethnicity by County and Mean CES Score",
    x = "County",
    y = "Proportion (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = "bottom"
  ) +
  scale_fill_manual(
  values = brewer.pal(
    n = length(unique(asthma_scores_plot$Race)),
    name = "Set3"
  )
)

```


### Dataset 3 Asthma Emergency Department Visits Dataset Description

This dataset contains information on asthma-related emergency department (ED) visits across counties, with a focus on demographic breakdowns by race/ethnicity and age group. The dataset has been filtered to include data only from the year 2019. Key variables in this dataset include the number of ED visits and an age-adjusted ED visit rate to account for age distribution differences across counties. The chosen variables provide a comprehensive view of ED utilization due to asthma across various demographic groups. By aggregating data by race/ethnicity and age group, we can explore potential disparities and patterns related to asthma.

### Dataset 3 Asthma ED Visits: Data Preparation for Analysis

```{r}
# Change to the lower case
asthma_ER_rate_county <- asthma_ER_rate_county %>%
  rename_with(~tolower(gsub(" ", "_", .x, fixed = TRUE)))

# Select Demographic Data of Interest
asthma_ed_visits_2019 <- asthma_ER_rate_county %>%
  filter(year == 2019, county != "California") %>%
  select(county, strata_name, age_group, number_of_ed_visits, `age_adjusted_ed_visit_rate`)

# Recode values that are reading in incorrectly
# Clean non-ASCII characters in 'strata_name' and identify 'age_group' and 'race_ethnicity'
asthma_ed_visits_2019 <- asthma_ed_visits_2019 %>%
  mutate(
    strata_name = str_replace_all(strata_name, "[^\x20-\x7E]", "-"),  # Replace non-ASCII characters
    age_group = case_when(
      strata_name %in% c("0-4 years", "5-17 years", "0-17 years", "18-64 years", "18+ years", "65+ years", "All ages") ~ strata_name,
      TRUE ~ NA_character_
    ),
    race_ethnicity = case_when(
      strata_name %in% c("White", "Black", "Hispanic", "Asian/PI", "AI/AN", "NHPI", "Multi-Race") ~ strata_name,
      TRUE ~ NA_character_
    )
  )

# Subset to most recent year and county level data
# Create a new variable 'age_category' based on the cleaned 'age_group'
asthma_ed_visits_2019 <- asthma_ed_visits_2019 %>%
  mutate(age_category = case_when(
    age_group %in% c("0-4 years", "5-17 years", "0-17 years") ~ "Children (0-17 years)",
    age_group %in% c("18-64 years", "18+ years") ~ "Adults (18-64 years)",
    age_group == "65+ years" ~ "Seniors (65+ years)",
    age_group == "All ages" ~ "All Ages",
    TRUE ~ "Unknown"
  ))

# Replace NA values in necessary columns
asthma_ed_visits_2019_clean <- asthma_ed_visits_2019 %>%
  mutate(
    age_group = replace_na(age_group, "Unknown"),
    race_ethnicity = replace_na(race_ethnicity, "Unknown"),
    age_category = replace_na(age_category, "Unknown")
  )

```

### Dataset 3 Asthma ED Visits: Table and Plot
```{r}
library(dplyr)
library(knitr)
library(kableExtra)

# Pivot table to only include one row per county
county_ed_visits_summary <- asthma_ed_visits_2019_clean %>%
  group_by(county) %>%
  summarize(
    total_ed_visits = round(sum(number_of_ed_visits, na.rm = TRUE), 2),
    mean_ed_visits = round(mean(number_of_ed_visits, na.rm = TRUE), 2),
    mean_age_adjusted_rate = round(mean(`age_adjusted_ed_visit_rate`, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  rename(
    County = county,
    `Total ED Visits` = total_ed_visits,
    `Mean ED Visits` = mean_ed_visits,
    `Mean Age-Adjusted Rate` = mean_age_adjusted_rate
  )

# Create Table
kable(county_ed_visits_summary, caption = "**Table: Total and Mean ED Visits, and Mean Age-Adjusted Rate per County**") %>%
  kable_styling(position = "center", full_width = FALSE)

```


```{r}
# County Data ED Visits (Descriptive Data)
library(dplyr)
library(knitr)
library(kableExtra)

# Create a new data frame summarizing total ED visits per county
county_ed_visits_plot_data <- asthma_ed_visits_2019_clean %>%
  group_by(county) %>%
  summarize(
    total_ed_visits = sum(number_of_ed_visits, na.rm = TRUE),
    .groups = "drop"
  )

print(county_ed_visits_plot_data)

# Create Plot
ggplot(county_ed_visits_plot_data, aes(x = reorder(county, -total_ed_visits), y = total_ed_visits)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Total ED Visits per County",
    x = "County",
    y = "Total ED Visits"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
# Descriptive Data Age Adjusted Rate
library(dplyr)
library(ggplot2)

# Create a new data frame summarizing mean age-adjusted ED visit rate per county
county_age_adjusted_rate_data <- asthma_ed_visits_2019_clean %>%
  group_by(county) %>%
  summarize(
    mean_age_adjusted_rate = mean(`age_adjusted_ed_visit_rate`, na.rm = TRUE),
    .groups = "drop"
  )

print(county_age_adjusted_rate_data)

# Plot mean age-adjusted ED visit rate per county
ggplot(county_age_adjusted_rate_data, aes(x = reorder(county, -mean_age_adjusted_rate), y = mean_age_adjusted_rate)) +
  geom_bar(stat = "identity", fill = "lightcoral") +
  labs(
    title = "Mean Age-Adjusted ED Visit Rate per County",
    x = "County",
    y = "Mean Age-Adjusted Rate"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels for readability
```

```{r}
# Pivot Data for Race/Ethnicity and Age Group
county_race_age_summary <- asthma_ed_visits_2019_clean %>%
  group_by(county, race_ethnicity, age_group) %>%
  summarize(
    total_ed_visits = sum(number_of_ed_visits, na.rm = TRUE),
    mean_ed_visits = mean(number_of_ed_visits, na.rm = TRUE),  # Mean number of visits per group
    mean_age_adjusted_rate = mean(`age_adjusted_ed_visit_rate`, na.rm = TRUE),
    .groups = "drop"
  )

# Display the summary table
library(knitr)
kable(
  county_race_age_summary,
  col.names = c("County", "Race/Ethnicity", "Age Group", "Total ED Visits", "Mean ED Visits", "Mean Age-Adjusted Rate"),
  caption = "Table: Summary of ED Visits by County, Race/Ethnicity, and Age Group"
)


```

#*BEGIN MILESTONE 4*
##Cleaning on Dataset 1 to prepare for joining with Dataset 3
```{r}

library(janitor)
library(dplyr)
library(stringr)

asthma_measures_df4_join <- asthma_measures_df2a_clean %>% 
  group_by(county) %>% 
  summarize(
    asthma_total_by_county_ds1 = sum(asthma, na.rm = TRUE),
    asthma_median_by_county_ds1 = median(asthma, na.rm = TRUE),
    pm2_5_median_by_county = median(pm2_5, na.rm = TRUE),
    diesel_pm_median_by_county = median(diesel_pm, na.rm = TRUE),
    traffic_median_by_county = median(traffic, na.rm = TRUE),
    tox_release_median_by_county = median(tox_release, na.rm = TRUE),
    housing_burden_median_by_county = median(housing_burden, na.rm = TRUE),
    education_median_by_county = median(education, na.rm = TRUE),
    unemployment_median_by_county = median(unemployment, na.rm = TRUE),
    poverty_median_by_county = median(poverty, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
ungroup() %>% 
  arrange(desc(asthma_median_by_county_ds1))

# Convert headers to snake_case and clean up county column
asthma_scores_demog %>%
  clean_names() %>% 
  mutate(county = str_remove(county, " County$"))


```

##Cleaning on Dataset 3 to prepare for joining with Dataset 1
```{r}
# Data Set 3 clean and standardize race/ethnicity categories
asthma_ed_join <- asthma_ed_visits_2019_clean %>%
  group_by(county) %>%
  summarize(
    # Total ED visits for the county
    asthma_total_by_county_ds3 = sum(number_of_ed_visits, na.rm = TRUE),
    asthma_median_by_county_ds3 = median(number_of_ed_visits, na.rm = TRUE),
    mean_age_adjusted_rate_ds3 = mean(age_adjusted_ed_visit_rate, na.rm = TRUE),
    median_age_adjusted_rate_ds3 = median(age_adjusted_ed_visit_rate, na.rm = TRUE),
    
    # Total and median visits by age groups
    total_children_visits = sum(ifelse(age_category == "Children (0-17 years)", number_of_ed_visits, 0), na.rm = TRUE),
    total_adults_visits = sum(ifelse(age_category == "Adults (18-64 years)", number_of_ed_visits, 0), na.rm = TRUE),
    total_seniors_visits = sum(ifelse(age_category == "Seniors (65+ years)", number_of_ed_visits, 0), na.rm = TRUE),
    median_children_visits = median(ifelse(age_category == "Children (0-17 years)", number_of_ed_visits, NA), na.rm = TRUE),
    median_adults_visits = median(ifelse(age_category == "Adults (18-64 years)", number_of_ed_visits, NA), na.rm = TRUE),
    median_seniors_visits = median(ifelse(age_category == "Seniors (65+ years)", number_of_ed_visits, NA), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(asthma_total_by_county_ds3))  # Sort in descending order by total ED visits

# View the combined table
print(asthma_ed_join)

```
## Join Datasets 1 and 3
```{r}

asthma_data_sets_join <- inner_join(asthma_measures_df4_join, asthma_ed_join, by = "county")

```

## Harmonize Dataset 2 County column data to prepare for joining
```{r}
library(janitor)
library(dplyr)
library(stringr)

# Convert headers to snake_case and clean up county column
asthma_scores_join <- asthma_scores_county %>%
  clean_names() %>% 
  mutate(county = str_remove(county, " County$"))
```

## Join All 3 Datasets
```{r}

# Join datasets 1,2,3 by county
asthma_all_join <- inner_join(asthma_data_sets_join, asthma_scores_join, by = "county")

# Join by datasets 1 and 2 by census tracts
asthma_measures_scores_subset_join <- inner_join(asthma_measures_df2a, asthma_scores_demog, by = "census_tract")

# Central Valley joined subset
asthma_central_valley_joined_subset  <- asthma_measures_scores_subset_join %>% 
  filter(county.x %in% c("Calaveras", "Fresno", "Kern", "Kings", "Madera", "Mariposa", "Merced", "San Joaquin", "Stanislaus", "Tulare", "Tuolumne")) 

```
## CalEnviron used 21 indicators to characterize Pollution Burden and Population Characteristics by County. Our study evaluated asthma, which is the Age Adjusted Rate of ED Visits, and its potential association with the select environmental pollution variables PM2.5, diesel PM, traffic, toxic release and the select population characteristic variables housing, education, unemployment and poverty. Results are shown below. 

## Scatterplots: Correlation between CalEnviron Variables and Median Age Adjusted Asthma ED Visit Rate by County

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(tidyr)

# Pivot to long dataset
asthma_long <- asthma_measures_df4_join %>%
  pivot_longer(
    cols = c(
      "pm2_5_median_by_county",
      "education_median_by_county",
      "unemployment_median_by_county",
      "poverty_median_by_county"
    ),
    names_to = "Variable",
    values_to = "Value"
  )

variable_labels <- c(
  "pm2_5_median_by_county" = "PM 2.5",
  "education_median_by_county" = "Education",
  "unemployment_median_by_county" = "Unemployment",
  "poverty_median_by_county" = "Poverty"
)

asthma_long$Variable <- factor(
  variable_labels[asthma_long$Variable],
  levels = variable_labels
)

# Create faceted scatterplot
ggplot(asthma_long, aes(x = Value, y = asthma_median_by_county_ds1)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  facet_wrap(~ Variable, scales = "free_x") +
  labs(
    title = "Median Asthma ED Visit Rates by County vs Selected Variables",
    x = "Variable Value",
    y = "Asthma Median by County"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 10),  # Adjust axis label font size
    plot.title = element_text(size = 12, hjust = 0.5)  # Center and resize title
  )
```
##Table: Correlation Coefficients between CalEnviron Variables and Median Age Adjusted Asthma ED Visit Rate by County
```{r}

install.packages("DT")
library(DT)


# Define the variables of interest

variables <- c(
  "pm2_5_median_by_county",
  "diesel_pm_median_by_county",
  "traffic_median_by_county",
  "tox_release_median_by_county",
  "housing_burden_median_by_county",
  "education_median_by_county",
  "unemployment_median_by_county",
  "poverty_median_by_county"
)
    
# Calculate the correlations and round to 3 decimal places

cor_results <- sapply(variables, function(var) {
  round(cor(asthma_measures_df4_join$asthma_median_by_county_ds1, asthma_measures_df4_join[[var]], use = "complete.obs"), 3)
})

# Create a data frame 
cor_df <- data.frame(
  Variable = c("PM2.5", "Diesel", "Traffic", "Toxic Release", "Housing Burden", "Education", "Unemployment", "Poverty"),  # Renamed variables
  Correlation = cor_results
)
# Rank the variables by the strength of the correlation (strongest to weakest)
cor_df <- cor_df[order(-cor_df$Correlation), ]  # Sort by Correlation in descending order

# Create interactive table 
datatable(
  cor_df,
  options = list(
    pageLength = 10,   # Display 10 rows per page
    autoWidth = TRUE,  # Automatically adjust column widths
    dom = 't'          # Hide the default table controls
  ),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: center; font-weight: bold; font-size: 16px;',
    "Correlation Coefficients for Median Asthma ED Visit Rates and Select Variables by County"
  )
)

```
##Table: Correlation Coefficients between Asthma ED Visit Rates and Select CalEnviron Variables by County Census Tracts

```{r}

# Define the variables of interest
variables <- c(
  "pm2_5",
  "diesel_pm",
  "traffic",
  "tox_release",
  "housing_burden",
  "education",
  "unemployment",
  "poverty"
)

cor_results <- sapply(variables, function(var) {
  round(cor(asthma_measures_df2a_clean$asthma, asthma_measures_df2a_clean[[var]], use = "complete.obs"), 3)
})

# Create a data frame 
cor_df <- data.frame(
  Variable = c("PM2.5", "Diesel", "Traffic", "Toxic Release", "Housing Burden", "Education", "Unemployment", "Poverty"),  # Renamed variables
  Correlation = cor_results
)
# Rank the variables by strength of the correlation (strongest to weakest)
cor_df <- cor_df[order(-cor_df$Correlation), ]  # Sort by Correlation in descending order

# Create interactive table 
datatable(
  cor_df,
  options = list(
    pageLength = 10,   
    autoWidth = TRUE, 
    dom = 't'          
  ),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: center; font-weight: bold; font-size: 16px;',
    "Correlation Coefficients for Asthma ED Visit Rates and Select Variables by County Census Tracts"
  )
)

```
### Boxplots:  Asthma ED Visit Rates by County Census Tracts
```{r}

library(ggplot2)
library(dplyr)
library(tidyr)

ggplot(asthma_measures_scores_subset_join, aes(x = reorder(county.x, asthma, FUN = median), y = asthma)) +
  geom_boxplot(
    fill = "lightblue", color = "darkblue", outlier.color = "red", outlier.size = 2
  ) +
  labs(
    title = "Asthma ED Visit Rates by County Census Tracts",
    x = "County",
    y = "Asthma"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
    axis.text.y = element_text(size = 5),  # Adjust size of county labels
    plot.margin = margin(3,,3,30)  # Add more space around the plot
  ) +
  coord_flip()


```
### Scatterplot: Mean CES 4.0 Score by County

```{r}

library(ggplot2)
library(dplyr)
library(tidyr)

ggplot(asthma_all_join, aes(x = reorder(county, asthma_median_by_county_ds1, FUN = median), y = mean_ces_score)) +
  geom_point(size = 2, color = "blue", alpha = 0.6) + 
  labs(
  title = "Mean CES Score by County",
  x = "County",
  y = "CES 4.0 Score"
  ) +
 theme_minimal() +
 theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
    axis.text.y = element_text(size = 5),  # Adjust size of county labels
    plot.margin = margin(3,,3,30)  # Add more space around the plot
  ) +
 coord_flip()

```
## Correlation between Median Age Adjsted Asthma ED Visits and Mean CES 4.0 Score by County

```{r}

install.packages("knitr")
library(knitr)
library(kableExtra)

# Calculate correlation coefficient
correlation <- cor(asthma_all_join$asthma_median_by_county_ds1, asthma_all_join$mean_ces_score, method = "pearson")

# Create a summary table with updated column names

correlation_table <- data.frame(
  Variable1 = "median asthma ED rate",
  Variable2 = "mean ces 4.0 score",
  Correlation = round(correlation, 3)
)

correlation_table %>%
  kable(align = c("l", "l", "c"), 
        caption = "Correlation Coefficient Between Median Age-Adjusted Asthma ED Visits and Mean CES 4.0 Score by County") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)

```
## Scatterplot: Correlation between Median Asthma ED Visit Rates and CES 4.0 Score
Correlation Coefficient 0.636

```{r}

library(ggplot2)
library(ggrepel) 

ggplot(asthma_all_join, aes(x = asthma_median_by_county_ds1, y = mean_ces_score)) +
  geom_point(size = 2, color = "blue", alpha = 0.6) + 
  geom_text_repel(
    aes(label = county),
    size = 2.5,
    max.overlaps = 25, 
    box.padding = 0.5,
    point.padding = 0.3
  ) +
  labs(
    title = "Correlation between Median Asthma Rates and CES 4.0 Score by County",
    subtitle = "Correlation Coefficient = 0.636",
    x = "Asthma",
    y = "CES Score"
  ) +
  theme_minimal()

```
##Boxplots: CES score vs Asthma ED Visit by County Census Tracts

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

asthma_scores_demog_subset <- asthma_scores_demog %>%
  select(county, census_tract, ces_4_0_score)%>%
  clean_names()%>%
  mutate(county = str_remove(county, " County$"))

asthma_measures_scores_subset_join <- inner_join(asthma_measures_df2a_clean, asthma_scores_demog_subset, by = "census_tract")
  

ggplot(asthma_measures_scores_subset_join, aes(x = reorder(county.x, asthma, FUN = median), y = ces_4_0_score)) +
  geom_boxplot(
    fill = "lightblue", color = "darkblue", outlier.color = "red", outlier.size = 2
  ) +
  labs(
    title = "CES Scores by County Census Tracts",
    x = "County",
    y = "CES Scores"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
    axis.text.y = element_text(size = 5),  # Adjust size of county labels
    plot.margin = margin(3,,3,30)  # Add more space around the plot
  ) +
  coord_flip()

```

## Correlation Coefficient Asthma ED Visits and CES 4.0 Score by County Census Tract

```{r}


# Calculate correlation coefficient
correlation <- cor(asthma_measures_scores_subset_join$asthma, asthma_measures_scores_subset_join$ces_4_0_score, method = "pearson")

correlation <- cor(asthma_measures_scores_subset_join$asthma, 
                   asthma_measures_scores_subset_join$ces_4_0_score, 
                   method = "pearson", 
                   use = "complete.obs")
print(correlation)

correlation_table <- data.frame(
  Variable1 = "asthma ED rate",
  Variable2 = "ces 4.0 score",
  Correlation = round(correlation, 3)
)

correlation_table %>%
  kable(align = c("l", "l", "c"), 
        caption = "Correlation Coefficient Between Asthma ED Visits and CES 4.0 Scores by Census Tracts") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)  

```
## Focus on Central Valley Counties. Eight out of the eleven Central Valley counties rank in the top 20 counties for highest ED visit rates by census tracts. 

## Below are Box Plots of ED Visit Rates by Central Valley Counties Census Tracts.

```{r}
asthma_central_valley_joined_subset  <- asthma_measures_scores_subset_join %>% 
  filter(county.x %in% c("Calaveras", "Fresno", "Kern", "Kings", "Madera", "Mariposa", "Merced", "San Joaquin", "Stanislaus", "Tulare", "Tuolumne")) 

ggplot(asthma_central_valley_joined_subset, aes(x = reorder(county.x, asthma, FUN = median), y = asthma)) +
  geom_boxplot(
    fill = "lightblue", color = "darkblue", outlier.color = "red", outlier.size = 2
  ) +
  labs(
    title = "Asthma ED Visits by Central Valley Counties' Census Tracts",
    x = "County",
    y = "Asthma"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
    axis.text.y = element_text(size = 5),  # Adjust size of county labels
    plot.margin = margin(3,,3,30)  # Add more space around the plot
  ) +
  coord_flip()

```
## ## Plotly Boxplots of ED Visit Rates by Central Valley Counties' Census Tracts

```{r}
library(plotly)
library(dplyr)

asthma_central_valley_joined_subseta <- asthma_central_valley_joined_subset %>%
  select(county.x, asthma, ces_4_0_score)


# Calculate medians and reorder county.x

asthma_central_valley_joined_subseta <- asthma_central_valley_joined_subseta %>%
  mutate(
    county.x = reorder(county.x, asthma, FUN = median, na.rm = TRUE)
  )

# Create the boxplot in ascending order of asthma medians while preserving census tract data

plot_ly(asthma_central_valley_joined_subseta,
        x = ~county.x,  # Reordered county.x by asthma median
        y = ~asthma,
        type = "box") %>%
  layout(
    title = "Asthma ED Visits by Central Valley Counties' Census Tracts",
    xaxis = list(title = "County"),
    yaxis = list(title = "Asthma ED Visits"),
    showlegend = FALSE
  )

```
## Boxplots of CES score by Central Valley Counties' Census Tracts

```{r}
ggplot(asthma_central_valley_joined_subset, aes(x = reorder(county.x, asthma, FUN = median), y = ces_4_0_score)) +
  geom_boxplot(
    fill = "lightblue", color = "darkblue", outlier.color = "red", outlier.size = 2
  ) +
  labs(
    title = "CES Scores by Central Valley Counties' Census Tracts",
    x = "County",
    y = "CES Scores"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
    axis.text.y = element_text(size = 5),  # Adjust size of county labels
    plot.margin = margin(3,,3,30)  # Add more space around the plot
  ) +
  coord_flip()

```
## Plotly Boxplots of CES Scores by Central Valley Counties' Census Tracts

```{r}

library(plotly)
library(dplyr)

asthma_central_valley_joined_subseta <- asthma_central_valley_joined_subset %>%
  select(county.x, asthma, ces_4_0_score)


# Calculate medians and reorder county.x

asthma_central_valley_joined_subseta <- asthma_central_valley_joined_subseta %>%
  mutate(
    county.x = reorder(county.x, asthma, FUN = median, na.rm = TRUE)
  )

# Create the boxplot in ascending order of asthma medians while preserving census tract data

plot_ly(asthma_central_valley_joined_subseta,
        x = ~county.x,  # Reordered county.x by asthma median
        y = ~ces_4_0_score,
        type = "box") %>%
  layout(
    title = "CES Scores by Central Valley Counties' Census Tracts",
    xaxis = list(title = "County"),
    yaxis = list(title = "CES Scores"),
    showlegend = FALSE
  )


```
###As per https://www.cdph.ca.gov/Programs/RPHO, Central California is defined as: 
###Calaveras, Fresno, Kern, Kings, Madera, Mariposa, Merced, San Joaquin, 
###Stanislaus, Tulare, Tuolumne


#Filtered Dataset for Central Valley Counties
```{r}

# Define the 11 Central Valley counties
central_valley_11 <- c("Calaveras", "Fresno", "Kern", 
                       "Kings", "Modera", "Mariposa", 
                       "Merced", "San Joaquin", "Stanislaus", 
                       "Sacramento", "San Joaquin", 
                       "Tulare", "Tuolumne")

# Filter the asthma_all_join dataset
asthma_central_valley_all_joined_dataset <- asthma_all_join %>%
  filter(county %in% central_valley_11)

# View the filtered dataset
print(asthma_central_valley_all_joined_dataset)

```

```{r}

library(dplyr)

environmental_data <- asthma_central_valley_all_joined_dataset %>%
  select(
    pm2_5_median_by_county,
    diesel_pm_median_by_county,
    traffic_median_by_county,
    tox_release_median_by_county,
    median_age_adjusted_rate_ds3
  )

# Calculate correlations
correlations_central_valley_env_median_age <- environmental_data %>%
  summarise(
    PM2.5 = cor(pm2_5_median_by_county, median_age_adjusted_rate_ds3, method = "spearman"),
    Diesel_PM = cor(diesel_pm_median_by_county, median_age_adjusted_rate_ds3, method = "spearman"),
    Traffic = cor(traffic_median_by_county, median_age_adjusted_rate_ds3, method = "spearman"),
    Toxic_Release = cor(tox_release_median_by_county, median_age_adjusted_rate_ds3, method = "spearman")
  )

# Print the Spearman correlations
print(correlations_central_valley_env_median_age)

```
# Correlation Coefficients Between Median Age-Adjusted ED Visit Rates and Age Groups in Central Valley, 2019
```{r}

library(ggplot2)
library(DT)  # For the interactive table
library(patchwork)

# Define the age groups
age_groups <- c(
  "median_children_visits",
  "median_adults_visits",
  "median_seniors_visits"
)

# Calculate the Spearman correlation for the table and save for plots
cor_results <- sapply(age_groups, function(var) {
  round(cor(asthma_central_valley_all_joined_dataset$median_age_adjusted_rate_ds3,
            asthma_central_valley_all_joined_dataset[[var]],
            use = "complete.obs", method = "spearman"), 3)
})

# Create a data frame for the table
cor_df <- data.frame(
  Age_Group = c("Children (0-17)", "Adults (18-64)", "Seniors (65+)"),
  Spearman_Correlation = cor_results
)

# Print correlation values for debugging
print(cor_df)

# Generate interactive table
datatable(
  cor_df,
  options = list(
    pageLength = 5,   # Display 5 rows per page
    autoWidth = TRUE,  # Automatically adjust column widths
    dom = 't'          # Hide the default table controls
  ),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: center; font-weight: bold; font-size: 16px;',
    "Spearman Correlation Coefficients for Median Age-Adjusted Rate and Age Groups"
  )
)

# Extract correlations for use in plots
cor_children <- cor_df$Spearman_Correlation[cor_df$Age_Group == "Children (0-17)"]
cor_adults <- cor_df$Spearman_Correlation[cor_df$Age_Group == "Adults (18-64)"]
cor_seniors <- cor_df$Spearman_Correlation[cor_df$Age_Group == "Seniors (65+)"]

# Scatter plot for children
scatter_children <- ggplot(asthma_central_valley_all_joined_dataset, aes(x = median_children_visits, y = median_age_adjusted_rate_ds3)) +
  geom_point(color = "black") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "black") +
  annotate(
    "text",
    x = quantile(asthma_central_valley_all_joined_dataset$median_children_visits, 0.8, na.rm = TRUE),
    y = quantile(asthma_central_valley_all_joined_dataset$median_age_adjusted_rate_ds3, 0.8, na.rm = TRUE),
    label = paste("r =", cor_children),
    size = 5,
    color = "black"
  ) +
  labs(
    title = "Children (0-17)",
    x = "Median Children Visits",
    y = "Median Age-Adjusted Rate"
  ) +
  theme_minimal()

# Scatter plot for adults
scatter_adults <- ggplot(asthma_central_valley_all_joined_dataset, aes(x = median_adults_visits, y = median_age_adjusted_rate_ds3)) +
  geom_point(color = "black") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "black") +
  annotate(
    "text",
    x = quantile(asthma_central_valley_all_joined_dataset$median_adults_visits, 0.8, na.rm = TRUE),
    y = quantile(asthma_central_valley_all_joined_dataset$median_age_adjusted_rate_ds3, 0.8, na.rm = TRUE),
    label = paste("r =", cor_adults),
    size = 5,
    color = "black"
  ) +
  labs(
    title = "Adults (18-64)",
    x = "Median Adults Visits",
    y = "Median Age-Adjusted Rate"
  ) +
  theme_minimal()

# Scatter plot for seniors
scatter_seniors <- ggplot(asthma_central_valley_all_joined_dataset, aes(x = median_seniors_visits, y = median_age_adjusted_rate_ds3)) +
  geom_point(color = "black") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "black") +
  annotate(
    "text",
    x = quantile(asthma_central_valley_all_joined_dataset$median_seniors_visits, 0.8, na.rm = TRUE),
    y = quantile(asthma_central_valley_all_joined_dataset$median_age_adjusted_rate_ds3, 0.8, na.rm = TRUE),
    label = paste("r =", cor_seniors),
    size = 5,
    color = "black"
  ) +
  labs(
    title = "Seniors (65+)",
    x = "Median Seniors Visits",
    y = "Median Age-Adjusted ED Visit Rate"
  ) +
  theme_minimal()

# Combine scatter plots
combined_plot <- scatter_children + scatter_adults + scatter_seniors + plot_layout(ncol = 3)

# Display the combined plot
print(combined_plot)
```

##Interpretation Spearman Correlation Coefficients for Median Age-Adjusted ED Visits Rate and Age Groups (Central Valley)

###The results shows a moderate positive correlation (0.465) for adults (18-64), indicating that changes in the studied factors moderately influence ED visit rates for this age group. For seniors (65+) and children (0-17), the weaker positive correlations (0.365 and 0.305, respectively) suggest a less pronounced relationship, but these groups are still affected to a lesser extent. Moving forward, we will explore which selected environmental factors correlate most strongly with each age group.

# Correlation Between Environmental Factors and Age Group Visits (Central Valley, 2019)
```{r}
# Load necessary libraries
# Load necessary libraries
library(dplyr)
library(DT)

# Create the correlation matrix as a data frame
correlation_matrix <- data.frame(
  Age_Group = c(
    "Children (0-17)", "Children (0-17)", "Children (0-17)", 
    "Adults (18-64)", "Adults (18-64)", "Adults (18-64)", 
    "Seniors (65+)", "Seniors (65+)", "Seniors (65+)"
  ),
  Environmental_Factor = c(
    "Diesel PM", "Traffic", "PM2.5",
    "Diesel PM", "Traffic", "PM2.5",
    "Diesel PM", "Traffic", "PM2.5"
  ),
  Correlation = c(
    0.6623991, 0.7357200, 0.3604946, 
    0.6273918, 0.8295413, 0.07018137, 
    0.6257399, 0.8144335, 0.06806552
  )
)

# Round the correlation values to 3 decimals and convert to character for consistent formatting
correlation_matrix <- correlation_matrix %>%
  mutate(Correlation = format(round(Correlation, 3), nsmall = 3)) %>%
  arrange(desc(as.numeric(Correlation))) # Sort by numeric value in descending order

# Display the sorted data frame as an interactive table
datatable(
  correlation_matrix,
  options = list(
    pageLength = 10,  # Display 10 rows per page
    autoWidth = TRUE, # Automatically adjust column widths
    dom = 't',        # Simplify controls (no search bar, just table)
    ordering = TRUE   # Enable sorting
  ),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: center; font-weight: bold;',
    "Correlation Between Environmental Factors and Age Group Visits (Central Valley, 2019)"
  )
)

```
```{r}
# Load required libraries
library(plotly)

# Create the dataset
correlation_matrix <- data.frame(
  Age_Group = c(
    "Children (0-17)", "Children (0-17)", "Children (0-17)", 
    "Adults (18-64)", "Adults (18-64)", "Adults (18-64)", 
    "Seniors (65+)", "Seniors (65+)", "Seniors (65+)"
  ),
  Environmental_Factor = c(
    "Diesel PM", "Traffic", "PM2.5",
    "Diesel PM", "Traffic", "PM2.5",
    "Diesel PM", "Traffic", "PM2.5"
  ),
  Correlation = c(
    0.6623991, 0.7357200, 0.3604946, 
    0.6273918, 0.8295413, 0.07018137, 
    0.6257399, 0.8144335, 0.06806552
  )
)

# Round correlation values to 3 decimals
correlation_matrix$Correlation <- round(correlation_matrix$Correlation, 3)

# Create the interactive scatter plot
scatter_plot <- plot_ly(
  data = correlation_matrix,
  x = ~Environmental_Factor,   # Environmental factors on the x-axis
  y = ~Correlation,            # Correlation values on the y-axis
  color = ~Age_Group,          # Color by age group
  type = 'scatter',            # Scatter plot type
  mode = 'markers',            # Use markers for points
  text = ~paste(
    "Age Group:", Age_Group, 
    "<br>Environmental Factor:", Environmental_Factor, 
    "<br>Correlation:", Correlation
  ),                           # Custom tooltip text
  hoverinfo = "text",          # Show only custom hover text
  marker = list(size = 15, opacity = 0.8) # Marker styling
) %>%
  layout(
    title = "Interactive Scatter Plot: Environmental Factors vs Age Group Correlations",
    xaxis = list(title = "Environmental Factor"),
    yaxis = list(title = "Correlation Coefficient"),
    legend = list(title = list(text = "Age Group")),
    hovermode = "closest"      # Display closest point on hover
  )

# Display the scatter plot
scatter_plot

```
## Interpretation
###The data from the Central Valley shows that traffic-related factors have the strongest correlations with ED visit rates across all age groups, especially for adults (18-64, 0.830) and seniors (65+, 0.814). Diesel PM exhibits moderate correlations, particularly for children (0-17, 0.662) and adults (18-64, 0.627), while PM2.5 shows weaker correlations across all groups.

## Correlation of Race Ethnicity and PM2.5 Measurements to Prepare for Geospatial Mapping
```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)

#set up the demographic variables
demog_var <- c("hispanic", "white", "black", "native_american", "asian", "other")

#calculate Pearson correlation coefficients
demog_cor_results <- sapply(demog_var, function(var) {
  round(cor(asthma_all_join[[var]], asthma_all_join$pm2_5_median_by_county, use = "complete.obs"), 3)
})

#make a data frame from my results
demog_cor_results <- data.frame(
  Demographic_Group = c("Hispanic", "White", "Black", "Native American", "Asian", "Other"),
  Correlation = sapply(
    c("hispanic", "white", "black", "native_american", "asian", "other"),
    function(var) round(cor(asthma_all_join[[var]], asthma_all_join$pm2_5_median_by_county, use = "complete.obs"), 3)
  )
)

#put them in rank order by absolute correlation
demog_cor_results <- demog_cor_results[order(-abs(demog_cor_results$Correlation)), ]

#make my kable table
library(knitr)
kable(
  demog_cor_results,
  caption = "Median PM2.5 Air Quality Measures Associated with Population Demography in California for Year 2019"
)

#make my plot
library(ggplot2)
ggplot(demog_cor_results, aes(x = reorder(Demographic_Group, Correlation), y = Correlation, fill = Correlation)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title = "Correlation Between Demographic Groups and PM2.5 Levels",
    x = "Demographic Group",
    y = "Pearson Correlation Coefficient"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )

```

##Prepare for Geospatial ER Asthma Choropleth Overlay Dot Density for Demography
```{r}
#get special packages especially shapefile and tigris for this niche work!
install.packages(c("sf", "tigris", "ggplot2", "dplyr"))
library(sf)
library(tigris)
library(ggplot2)
library(dplyr)
```

###Prepare the data from US shapefiles, and merge with master df via left join
```{r}
#bring in the U.S. county shapefile but retain sf format
options(tigris_use_cache = TRUE)
us_counties <- counties(cb = TRUE, class = "sf")

#filter the US counties to match those in asthma_all_join dataset
ca_counties <- us_counties %>% 
  filter(STATEFP == "06")  #note that "06" is the state code for CA

#do a left join of asthma data with CA geospatial data
asthma_geo <- ca_counties %>%
  left_join(asthma_all_join, by = c("NAME" = "county"))

```
##Prepare Centroids for Dot Density to supply the x and y needed for plots

```{r}
#pull in the centroids for county geometries
asthma_geo_centroids <- asthma_geo %>%
  st_centroid() %>%
  st_coordinates() %>%
  as.data.frame() %>%
  rename(x = X, y = Y) %>%
  bind_cols(asthma_geo)

#organize my demographic dot density data
asthma_demog_dots <- asthma_all_join %>%
  select(county, hispanic, white, black, native_american, asian, other) %>%
  tidyr::pivot_longer(
    cols = -county,
    names_to = "demographic_group",
    values_to = "population_proportion"
  ) %>%
  left_join(asthma_geo_centroids, by = c("county" = "NAME"))

```

##Plot the Choropleth ER by County with Dot Density Visual for Demography

```{r}
ggplot() +
  #make choropleth layer for asthma ER visit rates
  geom_sf(data = asthma_geo, aes(fill = median_age_adjusted_rate_ds3), color = NA) +
  scale_fill_gradient(low = "white", high = "red", name = "ER Visit Rate") +
  #make dot density layer for demographic characteristics
  geom_point(data = asthma_demog_dots, aes(
    x = x, y = y, size = population_proportion, color = demographic_group
  ), inherit.aes = FALSE, alpha = 0.6) +
  scale_color_brewer(palette = "Set3", name = "Population Demographic Group") +
  theme_minimal() +
  labs(
    title = "Geographic Map of ER Visit Rates with Demographic Density",
    subtitle = "Counties shaded by asthma-related emergency visit rates; dots represent demographic density",
    caption = "Data Source: California EnviroScore CES Data and US Census TIGRIS"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

```
#**Delete this line before submitting work: END Milestone 4**


# Data Dictionary: 
# Analysis of California Environmental Factors and Asthma Rates
## New Variables for Analysis:

##Terms used througout: "Central California" is defined as: Calaveras, Fresno, Kern, Kings, Madera, Mariposa, Merced, San Joaquin, Stanislaus, Tulare, Tuolumne; refer to https://www.cdph.ca.gov/Programs/RPHO.

## Dataset 1:
# asthma_median_by_county: median age-adjusted rate of emergency department visits for asthma, by county.
# education_median_by_county: median percent of population over 25 with less than a high school education, by county
# poverty_median_by_county: median percent of population living below two times the federal poverty level, by county
# unemployment_median_by_county: median percent of the population over the age of 16 that is unemployed and eligible for the labor force, by county
# housing_burden_median_by_county: median percent housing burdened low income households, by county
# pm2_5_median_by_county: median annual mean PM 2.5 concentrations, by county
# diesel_pm_median_by_county: median diesel PM emissions from on-road and non-road sources, by county
# traffic_median_by_county: median traffic density, in vehicle-kilometers per hour per road length, within 150 meters of the census tract boundary, by county
# tox_release_median_by_county: median toxicity-weighted concentrations of modeled chemical releases to air from facility emissions and off-site incineration (from RSEI)


## Dataset 2:
# county: California county name 
# mean_CES_score: mean CES score by county as a simple average
# total_population: sum population from census tracts aggregated by county
# hispanic: hispanic race proportion by county, a weighted avg of total_population
# white: White race proportion by county, a weighted avg of total_population
# black: Black race proportion by county, a weighted avg of total_population
# native_american: Native American race proportion by county, a weighted avg of total_population
# asian: Asian race proportion by county, a weighted avg of total_population
# other: Other race proportion by county, a weighted avg of total_population


## Dataset 3:
# county: The name of the county where the data is collected.
# strata_name: Original demographic categorization, includes both age and race/ethnicity information
# age_group: Age group of individuals visiting the ED
# race_ethnicity: Race/ethnicity of individuals visiting the ED
# age_category: Broader age categorization based on age_group
# number_of_ed_visits: Total number of asthma-related ED visits for the specified demographic in 2019
# age_adjusted_ed_visit_rate: Age-adjusted rate of ED visits per 100,000 population
# total_ed_visits: Aggregated count of ED visits per county 
# mean_ed_visits: Average number of ED visits per county or demographic subgroup
# mean_age_adjusted_rate: Average age-adjusted ED visit rate per county or demographic subgroup