---
title: "Milestone #2, Group 4"
author: "Avinia, Eve, Kelly"
<<<<<<< HEAD
date: "6 Oct 2024"
output: html_document format
=======
date:  '`6 Oct 2024`'
output: html_document 
>>>>>>> 6525781 (Changed html)
---

1.Description of dataset:

Q1a: What is the data source? (1-2 sentences on where the data is coming from, dates included, etc.)

Datasets 1 and 2 are both part of the California Communities 
Environmental Health Screening Tool 
(CalEnviroScreen), which was developed by the 
Office of Environmental Health Hazard Assessment, or # OEHHA, 
an office within the California Environmental Protection Agency.

This update to CalEnviroScreen, Version 4.0 was released in October 2021, and 
contains data from 2019. It is is an "evolving science-based method
for identifying impacted communities by taking into consideration pollution 
exposure and its effects, as well as health and socioeconomic status, 
at the census-tract level."

Dataset 3 Asthma Emergency Department Visits by County (2015-2022).
This dataset provides the number of asthma emergency department visits by county
from 2015 to 2022, sourced from the California Department of Health Care Access 
and Information.


# Q1b: How does the dataset relate to the group problem statement and question?

Dataset 1 includes 21 statewide environmental measures (indicators) which 
characterize both Pollution Burden (Exposures and Environmental Effects) and 
Population Characteristics (Sensitive Populations and Socioeconomic Factors). 
Asthma, in the dataset, is defined as one of the Sensitive Populations. 
This allows us to link all of the environmental measures to people with 
asthma, by county, age, race and ethnicity information. 
Dataset 1 will be useful to identify which specific types of interventions
could be undertaken to improve outcomes for people with asthma living in
counties disproportionately affected by air pollution.

Dataset 1 derives scores using percentiles for each of the 21 indicators in a 
given geographic area. Scores are averaged for the set of indicators in each
of the four components (Exposures, Environmental Effects, Sensitive Populations, 
and Socioeconomic Factors). Linking data from Dataset 1 to Dataset 3, 
which includes ED visits for asthma by county and demographics, 
may provide further evidence for targeted interventions.
 
Dataset 2 combines the four component scores to produce a CalEnviroScreen score
for a given place relative to other places in the state,


# 2. Import statements for all datasets:

# NOTE: Please use datasets available in the PHW251 Project Data github 
# repoLinks to an external site. 
# Use appropriate import function and package based on the type of file
# Utilize function arguments to control relevant components 
# (i.e. change column types, column names, missing values, etc.)
# Document the import process
# Clean up column names (recommend using snake case, which is lower case text with underscores between words)

```{r}

# Import dataset 1

library(tidyverse)
library(dplyr)
library(readr)

file_path_csv <- ("/home/rstudio/asthma_group4/data/calenviroscreen_measures_2021.csv")
asthma_measures <- read_csv(file_path_csv)
asthma_measures <- readr::read_csv(file_path_csv)
asthma_measures <- readr::read_csv(file_path_csv, col_names = TRUE, 
      col_types = NULL, skip = 0, skip_empty_rows = TRUE, na = "NA")
janitor :: clean_names(asthma_measures)

```

Kelly: alternative to the above, I needed to use different code to do this because the file path caused an error in my R studio set up

```{r}
# Import dataset 1, 2 and 3 (Avinia to confirm if dataset 3 OK)

library(tidyverse)
library(dplyr)
library(readr)

# Read Dataset 1 CSV with the specified arguments
asthma_measures <- readr::read_csv("data/calenviroscreen_measures_2021.csv", 
                            col_names = TRUE, 
                            col_types = NULL, 
                            skip = 0, 
                            skip_empty_rows = TRUE)


# Then clean the names using janitor
asthma_measures <- janitor::clean_names(asthma_measures)

# next work with Dataset 2...
# Read Dataset 2 CSV with the specified arguments
asthma_scores_demog <- readr::read_csv("data/calenviroscreen_scores_demog_2021.csv", 
                            col_names = TRUE, 
                            col_types = NULL, 
                            skip = 0, 
                            skip_empty_rows = TRUE)

# Then clean the names using janitor
asthma_scores_demog <- janitor::clean_names(asthma_scores_demog)

# next work with Dataset 3...
# Read Dataset 3 CSV with the specified arguments
asthma_ER_rate_county <- readr::read_csv("data/asthma-emergency-department-visit-rates-by-county-2015_2022.csv", 
                            col_names = TRUE, 
                            col_types = NULL, 
                            skip = 0, 
                            skip_empty_rows = TRUE)

# Then clean the names using janitor
asthma_ER_rate_county <- janitor::clean_names(asthma_ER_rate_county)

print(asthma_measures);print(asthma_scores_demog);print(asthma_ER_rate_county)
```


# 3. Identify data types for 5+ data elements/columns/variables:

# Identify 5+ data elements required for your specified scenario. 
# If <5 elements are required to complete the analysis, please choose 
# additional variables of interest in the data set to explore in this milestone.

# Utilize functions or resources in RStudio to determine the types of 
# each data element (i.e. character, numeric, factor)
# Identify the desired type/format for each variable—will you need to
# convert any columns to numeric or another type? 

Dataset 1 Columns to use of our 5+ variables of interest: 
California County, Asthma, Asthma Percentile, 
Diesel PM,  Diesel PM Percentile, Toxic Release,Toxic Release Percentile, 
Traffic, Traffic Percentile, PM2.5 and PM2.5 percentile, 
Education Percentile, Unemployment Percentile, 
Housing Burden Percentile, Poverty Percentile

Dataset 2 Columns to use for our 5+ variables of interest:
County, CES 4 score, CES 4 percentile, total population, children %, 
pop 10-64 % elderly %, hispanic %, white %, african %, native %, asian %

Dataset 3 Columns to use for our 5+ variables:
Number_of_ed_visits for asthma, Age_adjusted_ed_visit_rate, County name
(identifier), Strata.

# 4. Provide a basic description of the 5+ data elements:

# Numeric: mean, median, range
# Character: unique values/categories
# Or any other descriptives that will be useful to the analysis

Dataset 1, Columns and description of data:
California County is a character value. The remainder of the environmental
measures are all numeric values and are selected because they may be associated
with asthma disease burden. These numeric values do not need to be converted.

Asthma in dataset 1 is the age-adjusted rate of emergency department visits
for asthma. This data will be correlated with similar data from dataset 3 which
also includes ED visits by county, and both of these will be correlated with 
county level CalEnviro Screen (CES) scores from dataset 2.

Dataset 2, Columns and description of data:
County: this is Character (correlates to "California County" in Dataset 1 and
"County" in Dataset 3). CES 4 score, CES 4 percentile, and demographic data:
these are Numeric (e.g., population percentages).The numeric data in Dataset 2
will help examine potential relationships between environmental factors and 
asthma burden.

Dataset 3, Columns and description of data: 
Number_of_ed_visits for asthma (numeric), Age_adjusted_ed_visit_rate (numeric),
County name (character), Strata (character). 
This dataset provides the primary variable of interest—county-level 
asthma ED visits. It will be compared with environmental and 
socio-demographic factors to determine potential correlations.

Key connections between the datasets: County will correlate to "California 
County" in Dataset 1 and 3, and the age adjusted ER rates in Dataset 3 
will correlate to the three age categories in Dataset 2, and there is also 
a column in Dataset 3 called "strata" which has options for "Race/Ethnicity"
"Child vs Adult," and "Age groups" which will also be able to be connected 
to the columns I have selected out of Dataset 2. 

# 5. Describe cleaning that each data source may need:

# NOTE: There is no requirement for any data cleaning in this milestone, 
# but there may be cleaning you can do when you import your data! 
# Please list out the anticipated data cleaning needed. 
# Examples:
# Data elements that need to be converted to a different type
# Data elements that need cleaning or re-categorization
# Data elements that may be used for future joins (i.e. state, county)

Dataset 1 County data will need to be recategorized in order to merge with
dataset 3 because dataset 1 County data is organized by census tract and 
zip codes. Dataset 3 data is organized at the county level. 
Dataset 1 data is based on data collected in 2019, and dataset 3 contains
data from 2015 to 2022. We will need to filter or subset data to 2021.
Datasets 2 and 3 contain data on age, but age categories do not match and
will need to be re-categorized.

Dataset 2:
Some general cleaning across all datasets will be to standardize County names
to allow for proper joins. Find any missing data in critical columns and 
use code to deal with blanks. Also in particular for Dataset 2 we will need to
standardize demographic group names, age categories, and age-adjusted ER rates
to what is in Dataset 3, which has specific thresholds for age that don't 
align to Dataset 2. Also for the demographic % values these are in decimal
but if there are other % we need to join in Set 1 or 3, check if these need
conversion from decimals so all are being expressed in the same units.

Dataset 3:
Missing Values Action Needed: These columns contain missing values that 
may need to be addressed depending on the analysis. Possible actions include
removing rows with missing values if complete data is required.
Imputing missing values using techniques such as the mean or median of the
column to maintain the dataset’s integrity. 
Columns Affected:
Age Group. Action Needed: Group age categories to align across datasets
across datasets 2 and 3.
Reason: Re-categorizing age groups can simplify the analysis and make 
results easier to interpret.
Standardizing column names when we merge data. 
Columns Affected: All columns.
Reason: Standardized column names make the dataset easier to work with.
